\Large \bf Cyclone: High Availability for Persistent Key Value Stores

### 2017-08-03 paper/paper.tex
Updates.

The flashlog is written out in segments of configurable size (we use 128KB
segments). A segment buffer is prepared in memory (volatile DRAM) using the
layout shown in Figure~\ref{fig:flashlog_page}. We do not allow objects in the
flashlog to cross a 4KB boundary - linking multiple objects together with a
special flag encoded into the size if necessary. We flush log segments out to
the log file using asynchronous direct IO, and therefore we fill one log segment
buffer while keeping IO to another one outstanding to the flash drive. To avoid
having to do a synchronous metadata flush, we preallocate (using the
posix\_fallocate call) a gigabyte worth of zero filled disk pages at the end of
the log file whenever we hit its end.

@@

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions
using two-phase commit that also need to recover from failure used in systems
such as FARM~\cite{farm}.

\section{Evaluation}
\label{sec:evaluation}
We evaluate Cyclone on a 12 node x86 Xeon cluster connected via a 10 GigE
switch. Three of the machines are equipped with 1.6TB Intel DC P3600 SSDs and
4*10 GigE ports. The remaining nine machines do not have SSDs and have only one
10 GigE port, serving as clients for most of the experiments. We turn on jumbo
frame support in the 10 GigE switch to enable maximum use of batching in
Cyclone. As with other work~\cite{faast}, we use DRAM on the machines to proxy
for NVDIMMs where necessary - the persistent memory needed never exceeds 64 MB
regardless of the size of the key value store or second level log on flash. We
divide the evaluation into three parts. First, we evaluate Cyclone's performance
with a single level log as a pure software packet switch. Next, we evaluate
performance when adding a second level of log on flash. Finally, we evaluate
performance when integrated with Rocksdb~\cite{rocksdb} as an alternative to
Rocksdb's write ahead log. Unless otherwise mentioned, we use a 60 byte header
followed by an optional payload for experiments. We log both the header and
payload. In all cases the server echoes the received entry (header and payload)
back to the client. We fix the number of KV Store threads at 32 using at most 8
physcial logs with associated control plane cores for running RAFT instances.

Cyclone is built on top of DPDK to apply software packet switching techniques to
the log replication problem. We therefore begin by systematically evaluating
optimizations applied in Cyclone to replicate the NVM log in
Figure~\ref{fig:network_opts} - with no payload. The y-axis reports latency seen
at the client (which means two network round trips with replication). Using
TCP/IP to replicate a RAFT log tops out at around 30K entries/s.  Switching to
DPDK (the line marked +DPDK) improves the throughput by an order of magnitude to
around 500K entries/s. Using batching (the line marked +batching) improves the
performance further bringing us close to a million entries/s. Scaling
horizontally to 8 physical logs (+8 phy logs) improves performance to close to
2M entries/s. Finally using all 4 ethernet ports on the machine to replicate
entries improves performance considerably to 6M entries/s. In all, performance
improves by 200X over the TCP/IP single log baseline. Cyclone also considerably
improves the latency for replication, from close to 100us with TCP/IP to around
30us at peak throughput.

@@

There are two factors that can have significant impact on Cyclone's
performance. First, the number of replicas dictates the outgoing message rate
from the leader replica and therefore increasing the replication factor can
decrease Cyclone's performance. Figure~\ref{fig:replicas} shows the impact of
varying replica count. Using only a single replica cuts out a network round trip
and shows the best unloaded latency (10 us) and peak throughput (near 10M
entries/s). Adding replicas decreases the peak throughput down to around 2M
entries/s with 5 replicas. We note that a number of previous pieces of
work~\cite{faast, farm} use three replicas and therefore we focus on three
replicas for the replicated cases we consider below. The second factor that
dictates Cyclone's performance is the size of the log entry being
replicated. Figure~\ref{fig:payload} shows the effect of increasing the payload
size from zero to 512 bytes. Peak throughput drops from 6M entries/s to
approximately 2M entries/s. At this replication rate, the leader replica needs
to transmit data at approximately 30 Gbit/s. Coupled with the cost of network
headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
network line rate bottleneck at this point.

@@

We now turn our attention from the network component of Cyclone to the storage
one by adding the second level flashlog. We evaluate the impact of adding
block storage to our hitherto pure packet switching scenario in
Figure~\ref{fig:flashlog}. Batching entries from the top level NVDIMM log to the
second level flashlog is clearly beneficial as adding flash storage at the
second level has almost no impact on peak performance in Cyclone for small
entries. The situation however changes for larger
entries. Figure~\ref{fig:flashlog_512} shows that using a 512 byte payload has a
significant impact on peak throughput - it drops to approximately 350K
ops/sec. This corresponds to around 50K 4KB IOPS to the SSD to write out the
flashlog pages. The peak for the drive is 160K IOPS using a queue depth
(concurrency) of 128. With 32 application threads we expect a lower peak
throughput around 40K IOPS explaining our bottleneck at 50K IOPS.  It is
possible to tune our observed performance further by aligning the flush boundary
to increase the number of outstanding requests - we do not do so in this paper,
keeping Cyclone agnostic to the exact size of entry being replicated. A final
point about Figure~\ref{fig:flashlog_512} is that once we are past the storage
bottleneck the latency spike is dramatic and large enough to trigger Cyclone's
failure detector and repeated retries from the clients. There are - therefore -
no points on the ``knee'' of the curve as in the pure packet switched one-level
log case.

@@

The final dimension we evaluate is of ganged operations. The primary purpose of
ganged operations is to avoid the need for distributed transactions to
manipulate what is a single shared memory image and therefore we were most
concerned about unloaded latency given the complexity of synchronizing different
replication quorums as well executing our rendezvous protocol on multiple
cores. We therefore setup an experiment where a single client - reflecting the
unloaded case - made ganged requests to the replicas. We varied the number of
cores participating in the request from one to the full complement of 32
application cores. Figure~\ref{fig:ganged} shows the results both using a single
level log as well as a two level log. The primary takeaway is that unloaded
latency increases slowly as we increase the number of active cores - to around
40 us from the baseline of 20 us. There are two causes of this. First, there is
a quicker rate of increase from 1 to 8 active cores, the reason being that this
corresponds to an increase in the number of physical logs that must be
synchronized on. There is some amount of synchronization involved in the
userspace DPDK driver for access to common resources for accessing the NIC and
this affects the ability to simultaneously issue replication messages on all
quorums. The smaller contribution to increasing latency that continues past the
8 core case is due to cacheline pingponging when executing the rendevous of
Figure~\ref{fig:ganged_ops}. Both these sources of latency drift could be
corrected using replication quorums mapped to dedicated NICs and more scalable
rendezvous designs (such as with machine aware broadcast
trees~\cite{broadcast_tree}). However we deemed the complexity of such
optimizations unnecessary. The added latency for most cases is well under the
extra round trip delay (including replication) for all phases of a 2PC.

@@

We now evaluate Cyclone integrated with the Rocksdb persistent key value
store. Rocksdb is a complex industrial strength persistent key value store and
this means that it is accompanied by a complex array of performance tuning knobs
to get the best performance from flash. We were also aware that SSD performance
is slated for dramatic increases in the coming years with the introduction of
new memory technology such as 3DXPoint based Optane SSDs. To demonstrate that
Cyclone is future proof we eliminated flash media performance from the picture
as far as the key value store is concerned by placing all files for the key
value store (SSTables) on a RAMdisk, which presumably represents the limit in
performance for flash in the near future. All log files on secondary storage
however - both Rocksdb's own write ahead log and the alternative of Cyclone's
second level flashlog - are placed on the SSDs.

We evaluate two different request sizes: 8 byte keys with 8 byte values and 8
byte keys with 256 byte values. We run 8 physical logs on 8 cores and devote 32
cores to Rocksdb. Since we are interested in performance of the log, which is
only used for update requests, our workload is 100\% writes.

Before evaluating with Rocksdb, we measure the baseline performance of
replicating the log with Cyclone for the given request sizes - Rocksdb performs
a no-op. We note that in addition to the key and value, we are also logging
RocksDB specific request data such as operation type and the request header.
Figure~\ref{fig:kv_baseline} shows the baseline performance for the chosen
request sizes. With the smaller request size, Cyclone can conservatively sustain
close to a million requests a second at a latency of just under 25us. With the
larger request size, Cyclone can sustain around 350K requests a second, again at
a latency of just under 25 us. Armed with these baseline numbers we now examine
how well Cyclone performs with Rocksdb.

@@

Next, we consider the problem of supporting Rocksdb's write-batch operation that
atomically writes a set of key-value pairs into the KV store. In order to
perform the operation with Cyclone managing the log, the client must issue a
ganged operation across cores owning the keys in the write batch. This is in
contrast to baseline Rocksdb where the entire write batch can be sent to any
core. A key concern here was whether Cyclone would add any latency to the
operation due to the extra synchronization needed across replication quorums and
participating application cores. We examine the problem for the unloaded case
and small updates in Figure~\ref{fig:rocksdb_multi} for a single client with
increasing number of keys in the batch - till 32 keys that covers all
application cores. The line labeled Rocksdb is with no write ahead logging. We
note an increasing latency for this baseline indicating Rocksdb takes longer
with larger key batches. The existing option of Rocksdb/wal has considerably
larger latency. Cyclone does an effective job of cutting down on this latency
even as it needs to pay a price for synchronizing multiple quorums and
application cores making it somewhat slower than running Rocksdb with no logging
for batched writes. A distributed 2PC transaction still remains more expensive
just in terms of network delay than using Cyclone.

>>>>

The flashlog is written out in segments of configurable size (we use 128KB
segments). A segment buffer is prepared in memory (volatile DRAM) using the
layout shown in Figure~\ref{fig:flashlog_page}. We do not allow objects in the
flashlog to cross a 4KB boundary - linking multiple objects together with a
special flag encoded into the size if necessary. We flush log segments out to
the log file using asynchronous direct IO, and therefore we continue to fill log
segment buffers while keeping IO to previously filled buffers outstanding to the
flash drive. We allocate enough buffers to keep a maximum of 32 outstanding
requests to the SSD. To avoid having to do a synchronous metadata flush, we preallocate (using the
posix\_fallocate call) a gigabyte worth of zero filled disk pages at the end of
the log file whenever we hit its end.

@@

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions
using two-phase commit - especially when taking into account failure recovery -
as used in systems such as FARM~\cite{farm}.

\section{Evaluation}
\label{sec:evaluation}
We evaluate Cyclone on a 12 node x86 Xeon cluster connected via a 10 GigE
switch. Three of the machines are equipped with 1.6TB Intel DC P3600 SSDs and
4*10 GigE ports. The remaining nine machines do not have SSDs and have only one
10 GigE port, serving as clients for most of the experiments. We turn on jumbo
frame support in the 10 GigE switch to enable maximum use of batching in
Cyclone. As with other work~\cite{faast}, we use DRAM on the machines to proxy
for NVDIMMs where necessary - the persistent memory needed never exceeds 64 MB
regardless of the size of the key value store or second level log on flash. We
divide the evaluation into three parts.

First, we evaluate Cyclone's performance with a single level log as a pure
software packet switch. For this purpose, Cyclone uses a dummy server stub that
simply echoes the client request back to it. Next, we evaluate
performance with the dummy stub, but with the addition of the second level of log
on flash. Finally, we evaluate performance when integrated with
Rocksdb~\cite{rocksdb} as an alternative to Rocksdb's write ahead log.

Unless otherwise mentioned, we use a 60 byte header followed by an optional
payload for experiments. We log both the header and payload. Also, unless
otherwise mentioned we use 8 physical logs (and associated RAFT instances) each
mapped to a dedicated core. The remaining 32 cores are dedicated to the stub
server or Rocksdb, as the case may be.

We begin by systematically evaluating network stack related optimizations
applied in Cyclone to replicate the NVM log in Figure~\ref{fig:network_opts} -
with no payload. The y-axis reports latency seen at the client (which means two
network round trips with replication). Using TCP/IP to replicate a RAFT log tops
out at around 30K entries/s.  Switching to DPDK (the line marked +DPDK) improves
the throughput by an order of magnitude to around 500K entries/s. Using batching
(the line marked +batching) improves the performance further bringing us close
to a million entries/s. Scaling to 8 physical logs (+8 phy logs) improves
performance to close to 2M entries/s. Finally using all 4 ethernet ports on the
machine to replicate entries improves performance considerably to 6M
entries/s. In all, performance improves by 200X over the TCP/IP single log
baseline. Cyclone also considerably improves the latency for replication, from
close to 100us with TCP/IP to around 30us at peak throughput.

One can draw two important conclusions from
Figure~\ref{fig:network_opts}. First, one can indeed treat log replication as a
software packet switching problem \emph{provided} the persistent log can be held
in directly attached memory eliminating the block interface from the critical
path. Second, relieving the CPU from the overhead of data copies has a
significant positive effect on performance. Finally, using multiple physical
logs is essential to exploiting the concurrency available at the level of the
network interface card (even a single one!) that would otherwise go wasted due
to the serializing abstraction of a single log. Finally, we note that DPDKs raw
packet forwarding capability on that configuration is close to 12M packets per
second. We do not reach that mark due to the overhead of running the replication
protocol rather than a simple switching decision per packet. Although batching
goes some way in amortizing this overhead it cannot be eliminated entirely
because consensus protocols are more complex than a simple packet switching
decision.

@@

Both the replica count and payload size can have significant impact on Cyclone's
network performance. First, the number of replicas dictates the outgoing message
rate from the leader replica and therefore increasing the replication factor can
decrease Cyclone's performance. Figure~\ref{fig:replicas} shows the impact of
varying replica count. Using only a single replica cuts out a network round trip
and shows the best unloaded latency (10 us) and peak throughput (near 10M
entries/s). Adding replicas decreases the peak throughput down to around 2M
entries/s with 5 replicas. We note that a number of previous pieces of
work~\cite{faast, farm} use three replicas and therefore we focus on three
replicas for the replicated cases we consider below. The second factor that
dictates Cyclone's performance is the size of the log entry being
replicated. Figure~\ref{fig:payload} shows the effect of increasing the payload
size from zero to 512 bytes. Peak throughput drops from 6M entries/s to
approximately 2M entries/s. At this replication rate, the leader replica needs
to transmit data at approximately 30 Gbit/s. Coupled with the cost of network
headers all four 10 GigE ports are now saturated and therefore Cyclone hits the
network line rate bottleneck at this point. One can contrast this with the
previous experiment where the bottleneck to throughput was the cost of running
the consensus protocol for batches of packets.

@@

We now turn our attention from the network component of Cyclone to the storage
one by adding the second level flashlog. We evaluate the impact of adding the
flashlog in Figures~\ref{fig:flashlog} and \ref{fig:flashlog_512}. The benefit
of the second level flashlog is that it lets us limit the amount of needed
NVDIMM memory. The hypothesis was that batching the movement of data from the
NVM log to the flashlog would be sufficient to hide the latency of the block IO
device from network speed replication. The results confirm that in terms of
latency for the low to moderate load cases, our two level log arrangement is
effective at hiding the latency of the NVMe SSD. The picture however is
different for peak throughput. For small updates there is no impact on
throughput when we add the SSD. On the other hand 
Figure~\ref{fig:flashlog_512} shows that using a 512 byte payload has a
significant impact on peak throughput - it drops to approximately 350K
ops/sec. This corresponds to around 50K 4KB IOPS to the SSD to write out the
flashlog pages. This is in fact the expected IOPS limit with the 32 outstanding
requests to the drive that we maintain (Section~\ref{sec:storage}).
A final point about Figure~\ref{fig:flashlog_512} is that once we are past the
storage bottleneck the latency spike is dramatic and large enough to trigger
Cyclone's failure detector and repeated retries from the clients. There are -
therefore - no points on the ``knee'' of the curve as in the pure packet
switched one-level log case.

One can therefore conclude that using a small amount of directly attached NVM is
effective at hiding the latency of a background block storage device from the
criticial path of replication. Further, for small updates the throughput
bottleneck is the capability of the software to run the consensus protocol but
shifts to the secondary storage device for larger request sizes.

@@

We now evaluate ganged operations. The primary purpose of
ganged operations is to avoid the need for distributed transactions to
manipulate what is a single shared memory image and therefore we were most
concerned about unloaded latency given the complexity of synchronizing different
replication quorums as well executing our rendezvous protocol on multiple
cores. We therefore setup an experiment where a single client - reflecting the
unloaded case - made ganged requests to the replicas. We varied the number of
physical log participating in the request from one to the full complement of 8.
Figure~\ref{fig:ganged} shows the results. Unloaded
latency increases slowly as we increase the number of active physical logs - to
around 32 us from the baseline of 21 us. The reason for this increase is that
the DPDK userspace NIC driver pipelines request processing for communication
between the CPU core and the NIC. Simultaneous replicaton on multiple physical
logs represents a worst case for this arrangement as all eight cores try to
send replication messages to their replicas, adding about a microsecond of
serial latency each, to dispatch requests to the NIC. Regardless, the experiment
underlines the value of our ganged operation design over a distributed
transaction that would need 4 network hops for the two phase commit (and
replication) using an external co-ordinator, a minimum of 40us.

@@

We now evaluate Cyclone integrated with the Rocksdb persistent key value
store. Rocksdb is a complex industrial strength persistent key value store and
an important target for practical use of Cyclone.  Rocksdb is accompanied by a
complex array of performance tuning knobs to get the best performance from
flash. Designing LSM tree based key value stores to extract good performance
from flash and make best use of DRAM on the machine is an area of active
research~\cite{flodb, triad}. We wished to make sure that our conclusions would
apply even as Rocksdb performance improves with the integration of new
ideas. Keeping this in mind, we configured RocksDB to place all files for the
key value store (SSTables) on a RAMdisk, which presumably represents the limit
in performance for both software enhancements as well as improvements to
flash-based SSDs. However - both Rocksdb's own write ahead log and the
alternative of Cyclone's second level flashlog - are placed on the SSDs. In
effect we therefore make availability a harder problem in this setting,
guaranteeing our design can handle the immediate future.

We use 8 bytes keys with either 8 bytes values or 256 byte values.
Since our focus is replication and only update requests are replicated, we use a
pure update workload to test the capability of the system.

Before evaluating with Rocksdb, we measure the baseline performance of
replicating the log with Cyclone for the given request sizes - Rocksdb performs
a no-op. We note that in addition to the key and value, we are also logging
RocksDB specific request data such as operation type and the request header,
making this different from the previous experiments.
Figure~\ref{fig:kv_baseline} shows the baseline performance for the chosen
request sizes. With the smaller request size, Cyclone can conservatively sustain
close to a million requests a second at a latency of just under 25us. With the
larger request size, Cyclone can sustain around 350K requests a second, again at
a latency of just under 25 us. Armed with these baseline numbers we now examine
how well Cyclone performs with Rocksdb.

@@

Next, we consider the problem of supporting Rocksdb's write-batch operation that
atomically writes a set of key-value pairs into the KV store. This is a
practical application of ganged operations in Cyclone. In order to
perform the operation with Cyclone managing the log, the client must issue a
ganged operation across physical logs owning the keys in the write batch.
A key concern here was whether Cyclone would add any latency to the
operation due to the extra synchronization needed across replication quorums (
shown in a previous experiment).
We examine the problem for the unloaded case
and small updates in Figure~\ref{fig:rocksdb_multi} for a single client with
increasing number of keys in the batch - till 32 keys.
The line labeled Rocksdb is with no write ahead logging. We
note an increasing latency for this baseline indicating Rocksdb itself takes longer
with larger key batches. The existing option of Rocksdb/wal has considerably
larger latency. Cyclone does an effective job of cutting down on this latency
even as it needs to pay a price for synchronizing multiple physical
logs. Cyclone therefore provides effective replicated logging for batched
operations using its idea of ganged operations, while scaling the performance of
single operations using its multiple physical logs.
